Plug-and-play dual-tree algorithm runtime analysis

### 2014-12-29 bounds-files/general-runtime-bound.tex
updated abstract and intro, typo fixes, added alternative imbalance notes

Next, we develop notions with which to understand the behavior of the cover tree
dual-tree traversla when the datasets are of significantly different scale
distributions.

If the datasets are similar in scale distribution (that is, inter-point distances
tend to follow the same distribution), then the recursion will alternate between
query recursions and reference recursions.  But if the query set contains points
which are, in general, much further apart than the reference set, then the
recursion will start with many query recursions before reaching a reference
recursion.  The converse case also holds.  We are interested in formalizing
this notion of scale distribution; therefore, define the following
dataset-dependent constants for the query set $S_q$ and the reference set $S_r$:

@@

\begin{equation}
\theta = \max\left(\min\left(O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
O(N^2)\right), \; 0\right).
\end{equation}
\end{lemma}

>>>>

Next, we develop notions with which to understand the behavior of the cover tree
dual-tree traversal when the datasets are of significantly different scale
distributions.

If the datasets are similar in scale distribution (that is, inter-point distances
tend to follow the same distribution), then the recursion will alternate between
query recursions and reference recursions.  But if the query set contains points
which are, in general, much farther apart than the reference set, then the
recursion will start with many query recursions before reaching a reference
recursion.  The converse case also holds.  We are interested in formalizing
this notion of scale distribution; therefore, define the following
dataset-dependent constants for the query set $S_q$ and the reference set $S_r$:

@@

\begin{equation}
\theta = \max\left\{\min\left[O(N \lceil \log_2(\delta_q / \delta_r) \rceil),
O(N^2)\right], \; 0\right\}.
\end{equation}
\end{lemma}

### 2014-12-29 bounds.tex
updated abstract and intro, typo fixes, added alternative imbalance notes

\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
Numerous machine learning algorithms contain pairwise statistical problems at
their core. Often, tree structures are used to solve these problems efficiently.
Dual-tree algorithms using cover trees have rigorous worst-case runtime
guarantees for some of these problems. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
the cover tree, separating out the problem-dependent and the problem-independent
elements. This allows us to just plug in bounds for the problem-dependent
elements to get runtime guarantees for dual-tree algorithms for any pairwise
statistical problem. We demonstrate this plug-and-play procedure for
nearest-neighbor search and approximate kernel density estimation to get
improved runtime guarantees.  Under some assumptions, we also present the first
linear runtime guarantee for dual-tree based range search.
\end{abstract}

@@

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems.
Nearest neighbor search is one such problem, naively requiring $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, naively requiring $O(N^2)$
time to answer $O(N)$ queries. The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

@@

Dual-tree algorithms can be easily understood through the recent framework of
\citet{curtin2013tree}: two trees (a query tree and a reference
tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
visits combinations of nodes from the trees in some sequence (each combination
consisting of a query node and a reference node), calling a problem-specific
\texttt{Score()} function to determine if the node combination can be pruned.
If not, then a problem-specific \texttt{BaseCase()} function is called for each
combination of points held in the query node and reference node.  This has
significant similarity to the more common single-tree branch-and-bound
algorithms, except that the algroithm must recurse into child nodes of {\it
both} the query tree and reference tree.

@@

Section \ref{sec:preliminaries} lays out the required background, notation, and
introduces the cover tree and its associated theoretical properties.  Readers
familiar with the cover tree literature and dual-tree algorithms
\citep[especially][]{curtin2013tree} may find that section to be review.
Following that, we introduce an intuitive measure of cover tree imbalance in
Section \ref{sec:imbalance}.  This measure of imbalance is then used to prove
the main result of the paper in Section \ref{sec:bound}, which is a worst-case
runtime bound for generalized dual-tree algorithms.  We apply this result to
three specific problems: nearest neighbor search (Section \ref{sec:nns}),
approximate kernel density estimation (Section \ref{sec:akde}), and range search
/ range count (Section \ref{sec:rs}), showing linear runtime bounds for each of
those algorithms.  Each of these bounds is an improvement on the
state-of-the-art, and in the case of range search, is the first such bound.
%  We
%then investigate the empirical performance of cover trees for these algorithms
%in Section \ref{sec:emp}, with a focus on understanding the dataset-dependent
%constants that our bounds use.

>>>>

\begin{abstract}
%A large collection of machine learning algorithms contain subproblems that can
%be efficiently solved with trees.  In the situation where a dual-tree algorithm
%with cover trees is used to solve a problem, rigorous worst-case runtime bounds
%can often be shown.  A worst-case runtime bound is proven for all cover tree
%dual-tree algorithms; this bound can be employed to give useful runtime bounds
%for individual problems.  We show a new, tighter bound of $O(c^9 \nu N)$ for
%all-nearest-neighbor search and also provide linear runtime bounds for
%approximate kernel density estimation and exact range count/search under some
%mild assumptions.  These are the first linear runtime bounds we are aware of for
%range count/search.
%Numerous machine learning algorithms contain pairwise statistical problems at
%their core. Often, tree structures are used to solve these problems efficiently.
%Dual-tree algorithms using cover trees have rigorous worst-case runtime
%guarantees for some of these problems. In this paper, we present a {\em
%problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
%the cover tree, separating out the problem-dependent and the problem-independent
%elements. This allows us to just plug in bounds for the problem-dependent
%elements to get runtime guarantees for dual-tree algorithms for any pairwise
%statistical problem. We demonstrate this plug-and-play procedure for
%nearest-neighbor search and approximate kernel density estimation to get
%improved runtime guarantees.  Under some assumptions, we also present the first
%linear runtime guarantee for dual-tree based range search.
Numerous machine learning algorithms contain pairwise statistical problems at
their core -- i.e. tasks that require computations over all pairs of input points if implemented directly. 
Often, tree structures are used to solve these problems efficiently. Dual-tree algorithms 
can efficiently solve or approximate many of these problems.
Using cover trees, rigorous worst-case runtime
guarantees have been proven for some of these algorithms. In this paper, we present a {\em
problem-independent} runtime guarantee for {\em any} dual-tree algorithm using
the cover tree, separating out the problem-dependent and the problem-independent
elements. This allows us to plug in bounds for the problem-dependent
elements to get runtime guarantees for dual-tree algorithms for any pairwise
statistical problem without re-deriving the entire proof. We demonstrate this plug-and-play procedure for
nearest-neighbor search and approximate kernel density estimation to get
improved runtime guarantees.  Under mild assumptions, we also present the first
linear runtime guarantee for dual-tree based range search.
\end{abstract}

@@

A surprising number of machine learning algorithms have computational
bottlenecks that can be expressed as pairwise statistical problems. 
By this, we mean computational tasks that can be evaluated directly by iterating over all 
pairs of input points. 
Nearest neighbor search is one such problem, since for every query point, we can 
evaluate its distance to every reference point and keep the closest one. 
This naively requires $O(N)$ time
to answer in single query in a reference set of size $N$; answering $O(N)$
queries subsequently requires prohibitive $O(N^2)$ time. Kernel density
estimation is also a pairwise statistical problem, since we compute a sum over all 
reference points for each query point.
This again requires $O(N^2)$
time to answer $O(N)$ queries if done directly. 
The reference set is typically indexed with
spatial data structures to accelerate this type of computation
\citep{finkel1974quad, langford2006}; these result in $O(\log N)$ runtime per
query under favorable conditions.

@@

Dual-tree algorithms can be easily understood through the recent framework of
\citet{curtin2013tree}: two trees (a query tree and a reference
tree) are traversed by a {\it pruning dual-tree traversal}.  This traversal
visits combinations of nodes from the trees in some sequence (each combination
consisting of a query node and a reference node), calling a problem-specific
\texttt{Score()} function to determine if the node combination can be pruned.
If not, then a problem-specific \texttt{BaseCase()} function is called for each
combination of points held in the query node and reference node.  This has
significant similarity to the more common single-tree branch-and-bound
algorithms, except that the algorithm must recurse into child nodes of {\it
both} the query tree and reference tree.

@@

Section \ref{sec:preliminaries} lays out the required background, notation, and
introduces the cover tree and its associated theoretical properties.  Readers
familiar with the cover tree literature and dual-tree algorithms
\citep[especially][]{curtin2013tree} may find that section to be review.
Following that, we introduce an intuitive measure of cover tree imbalance, an 
important property for understanding the runtime of dual-tree algorithms, in
Section \ref{sec:imbalance}.  This measure of imbalance is then used to prove
the main result of the paper in Section \ref{sec:bound}, which is a worst-case
runtime bound for generalized dual-tree algorithms.  We apply this result to
three specific problems: nearest neighbor search (Section \ref{sec:nns}),
approximate kernel density estimation (Section \ref{sec:akde}), and range search
/ range count (Section \ref{sec:rs}), showing linear runtime bounds for each of
those algorithms.  Each of these bounds is an improvement on the
state-of-the-art, and in the case of range search, is the first such bound.
%  We
%then investigate the empirical performance of cover trees for these algorithms
%in Section \ref{sec:emp}, with a focus on understanding the dataset-dependent
%constants that our bounds use.
